Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-03 paper/main.tex
minor
abstract

\abstract{\textbf{Motivation:} Current state of the art gene annotation programs for eukaryotic genomes have to be pretrained for each species and still often produce annotations that lack the required accuracy. \\
\textbf{Results:} We set up a framework that can be used to build much more general gene callers that also outperform traditional methods on a variety of complex genomes.\\
\textbf{Availability:} \\
\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\begin{figure*}[!tpb]
\label{fig:main_results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results and other investigations. The top row shows the animal data; the bottom row is concerned with the plants. Yellow dots are for AUGUSTUS and the blue ones are for Helixer. {\bf a)} Swarm - and boxplots showing the Subgenic F1 scores for all genomes that were not used during training of Helixer, which is equivalent to all validation and test genomes. Each dot represents the prediction performace for one genome. The median prediction scores are shown by the red line. Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf b)} Scatterplots showing the prediction performances measured in Subgenic F1 by genome length. The regression line is shown with a 95\% confidence interval. {\bf c)} The average Genic F1 and basepair wise accuracy with respect to the position in the 20000 basepair long input sequences. The dashed line shows the same without overlapping. Each value is the average performance of a 200 basepair long subsection by one of the models of the final ensemble. The usage of the model without overlapping results in a significant negative bias towards the borders of each sequence. Overlapping the predictions helps to mitigate a lot of that while showing no negative side effects elsewhere on average. See Section 4 for the Supplements for the effects of overlapping for each individual species.}
\end{figure*}

@@

>>>>

\abstract{\textbf{Motivation:} Current state-of-the-art tools for the annotation of genes in eukaryotic genomes have to be pretrained for each species and still often produce annotations that can be improved much further. The fundamental algorithmic architecture for those tools also remains the same since about two decades, limiting learning capabilities. Here, we set out to improve the cross-species annotation of genes from DNA data alone with the help of deep learning. The goal is to eliminate the dependency on a closely related gene model while also improving the predictive quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of a cross-species deep learning model that improves significantly on performance and generalizability when compared to more traditional methods. We evaluate our approach by building a single model for the annotation of 186 animal genomes as well as one for 51 plants. Our predictions are shown to be much less sensitive to the length of the genome than those of a current state-of-the-are tool. We also present two novel post-processing techniques that each worked to further strengthen our annotations and show in-depth results of an RNA-Seq based comparison of our predictions.\\
\textbf{Availability:} The source code of this work is available at \url{https://github.com/weberlab-hhu/HelixerPrep} [change] under the GNU General Public License v3.0.\\
\textbf{Contact:} \href{alisandra.denton@hhu.de}{alisandra.denton@hhu.de}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\begin{figure*}[!tpb]
\label{fig:main_results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results and other investigations. The top row shows the animal data; the bottom row is concerned with the plants. Yellow dots are for AUGUSTUS and the blue ones are for Helixer. {\bf a)} Swarm - and boxplots showing the Subgenic F1 scores for all genomes that were not used during training of Helixer, which is equivalent to all validation and test genomes. Each dot represents the prediction performace for one genome. The median prediction scores are shown by the red line. Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf b)} Scatterplots showing the prediction performances measured in Subgenic F1 by genome length. The regression line is shown with a 95\% confidence interval. {\bf c)} The average Genic F1 (in red) and basepair wise accuracy (in purple) with respect to the position in the 20000 basepair long input sequences. The dashed line shows the same without overlapping. Each value is the average performance of a 200 basepair long subsection by one of the models of the final ensemble. The usage of the model without overlapping results in a significant negative bias towards the borders of each sequence. Overlapping the predictions helps to mitigate a lot of that while showing no negative side effects elsewhere on average. See Section 4 for the Supplements for the effects of overlapping for each individual species.}
\end{figure*}

@@

Eliminating the need for a close genome annotation being present for pretraining.
